First Order Alternation

### 2018-10-12 quantifiers/draft.tex
ok

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we denote by $X
\mapsto \Data$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols} $(a,\nu)$, where $a$ is an input symbol and
$\nu$ is a valuation. A \emph{data word} (simply called word in the
following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

A \emph{first-order alternating automaton} (automaton) is a tuple $\A
= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
\item $\Sigma$ is the finite set of input events, 
\item $X$ is a finite set of input variables, 
\item $Q$ is a finite set of predicates denoting control states, 
\item $\iota \in \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
\item $F \subseteq Q$ is the set of predicates denoting final states, and 
\item $\Delta$ is a set of \emph{transition rules} of the form
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
  a predicate, $a \in \Sigma$ is an input event and $\psi \in
  \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive
  formula, where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.
\end{compactitem}
The quantifiers occurring in the right-hand side formula of a
transition rule are referred to as \emph{transition quantifiers}. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
other than predicate symbols, such that the labeling is compatible
with the transition relation of the automaton (point
\ref{it3:unfolding} of Definition \ref{def:unfolding}).

Suppose that a given input event sequence $\alpha \in \dom(U)$ is
labeled with a sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1),
\ldots U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement
from Definition \ref{def:unfolding} (\ref{it3:unfolding}) and,
moreover, it has a non-local GLI $(I_0, \ldots, I_n)$, such that
$\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
renamed variables occurring in the rules of $\Delta$ under the scope
of existential transition quantifiers (formally these are the
existentially quantified variables from the prenex normal form of
$\quantform{\alpha_k}$). The interpolant is used in refining the
labeling of the sequence as follows, for all $k \in [n]$: \begin{compactitem}
\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
  from $I_k$ by\begin{inparaenum}[(i)]
    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
      %
    \item existentially quantifying each variable $y \in
      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$.
  \end{inparaenum}
%
\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
\end{compactitem}

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of\begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data
values seen in the input at some earlier moment in time, or 
%
\item variables $\set{x_{i_1},\ldots,x_{i_k}}$ that track past choices
  made within the transition rules. 
\end{inparaenum}
From the present point of view, it is not important when exactly in
the past a certain input has been read or when a choice has been made,
but rather the (symbolic) value and its associated data
constraints. Intuitively, existential quantification of these
variables does the job of ignoring the particular moments in the past
when they have been encountered. Formally, to see that the new set of
formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
\ref{def:unfolding}, it is enough to notice the following
facts: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}.} $\stamp{\iota}{0} \models
  I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) = U(\alpha_0)
  \wedge J_0 \equiv \iota \wedge J_0$.
%
\item by Definition \ref{def:generalized-lyndon-interpolant}
  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
  [n-1]$:
  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
  \leftarrow \stamp{\psi}{k+1} \models I_{k+1}\] By slightly abusing
  notation, let $\stamp{I}{n}_\ell$ be the formula in which every
  predicate symbol $\stamp{q}{\ell}$ has been replaced with
  $\stamp{q}{n}$. Then the following entailment holds: 
  \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \leftarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
  Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \leftarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
  satisfies the labeling condition of Definition \ref{def:unfolding}
  (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
  we obtain, as required: 
  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \leftarrow \stamp{\psi}{1} \models
  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
\end{compactitem}

@@

Nevertheless, if we require just a yes/no answer (i.e.\ not an
interpolant) recently developped quantifier instantiation heuristics
\cite{ReynoldsKK17} perform rather well in answering a large number of
queries in this class. Observe, moreover, that coverage does not need
to rely on a complete decision procedure. If the prover fails in
answering the above satisfiability query, then the semi-algorithm
assumes the node is not covered and continues exploring its
successors. Failure to compute complete coverage may lead to
divergence (non-termination) and failure to prove emptiness, but the
soundness of the procedure is not affected.

>>>>

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we
denote by $X \mapsto \Data$ the set of valuations of the variables $X$
and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
symbol and $\nu$ is a valuation. A \emph{data word} (simply called
word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$. A
\emph{first-order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
\item $\Sigma$ is a finite set of input events, 
\item $X$ is a finite set of input variables, 
\item $Q$ is a finite set of predicates denoting control states, 
\item $\iota \in \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
\item $F \subseteq Q$ is the set of predicates denoting final states, and 
\item $\Delta$ is a set of \emph{transition rules} of the form
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
  a predicate, $a \in \Sigma$ is an input event and $\psi \in
  \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive
  formula, where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.
\end{compactitem}
The quantifiers occurring in the right-hand side formula of a
transition rule are referred to as \emph{transition quantifiers}. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
other than predicate symbols, such that the labeling is compatible
with the transition relation of the automaton, according to the point
(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. Suppose that
a given input event sequence $\alpha \in \dom(U)$ is labeled with a
sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1), \ldots
U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement from
Definition \ref{def:unfolding} (\ref{it3:unfolding}) and, moreover, it
has a non-local GLI $(I_0, \ldots, I_n)$, such that $\voc{I_k}
\subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
renamed variables occurring in the rules of $\Delta$ under the scope
of existential transition quantifiers\footnote{These are the
  existentially quantified variables from the prenex normal form of
  $\quantform{\alpha_k}$.}. The interpolant is used to refine the
labeling of the sequence as follows, for all $k \in
[n]$: \begin{compactitem}
\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
  from $I_k$ by \begin{inparaenum}[(i)]
    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
      %
    \item existentially quantifying each variable $y \in
      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$,
  \end{inparaenum}
%
\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
\end{compactitem}

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data
values seen in the input at some earlier moment in time, or 
%
\item variables $\set{x_{i_1},\ldots,x_{i_k}}$ that track past choices
  made within the transition rules. 
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, as only the value
of the variable determines the future behavior. Intuitively,
existential quantification of these variables does the job of ignoring
when in the past these values have been seen. Formally, the new set of
formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
\ref{def:unfolding}, because: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, since, by point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
  \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
  U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
%
\item by Definition \ref{def:generalized-lyndon-interpolant}
  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
  [n-1]$:
  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
  \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] Then the following
  entailment holds:
  \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
  Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
  satisfies the labeling condition of Definition \ref{def:unfolding}
  (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
  we obtain, as required: 
  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models
  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
\end{compactitem}

@@

Nevertheless, if we require just a yes/no answer (i.e.\ not an
interpolant) recently developped quantifier instantiation heuristics
\cite{ReynoldsKK17} perform rather well in answering a large number of
queries in this class. Observe, moreover, that coverage does not need
to rely on a complete decision procedure. If the prover fails in
answering the above satisfiability query, then the semi-algorithm
assumes the node is not covered and continues exploring its
successors. Failure to compute complete coverage may lead to
divergence (non-termination) and failure to prove emptiness, but does
not affect the soundness of the semi-algorithm (real counterexamples
will still be found).

\section{Applications}
\label{sec:applications}

The main application of first-order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, we can extend our previous
\emph{generic register automata} \cite{HolikIosifRogalewiczVojnar18}
inclusion checking framework, by allowing monitor (right-hand side)
automata to have local variables, that are not visible in the
language\footnote{We conjecture that hidden computations using
  invisible variables (clocks) are the reason for which timed and
  register automata cannot be complemented within their class.}.

Another application is checking safety (mutual exclusion, absence of
deadlocks, etc.) and liveness (termination, lack of starvation, etc.)
properties of parameterized concurrent programs, consisting of an
unbounded number of replicated threads that communicate via a fixed
set of global variables (locks, counters, etc.). The verification of
parametric programs has been reduced to checking the emptiness of a
(possibly infinite) sequence of first-order alternating automata,
called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
the inclusion of the set of traces of a parametric concurrent program
into increasingly general \emph{proof spaces}, obtained by
generalization of counterexamples. The program and the proof spaces
are first-order alternating automata over the infinite alphabet of
pairs consisting of program statements and thread identifiers.

\subsection{Timed Automata}
\label{sec:ta}

The standard definition of a finite \emph{timed word} is a sequence of
pairs $(a_1,\tau_1), \ldots, (a_n,\tau_n) \in (\Sigma \times
\real)^*$, where $\real$ is the set of real numbers, such that $0 \leq
\tau_i < \tau_{i+1}$, for all $i \in [1,n-1]$. Intuitively, $\tau_i$
is the moment in time where the input event $a_i$ occurs. Given a set
$C$ of \emph{clocks}, the set $\Phi(C)$ of \emph{clock constraints} is
defined inductively as the set of formulae $x \leq c$, $x \geq c$,
$\neg\delta$, $\delta_1 \wedge \delta_2$, where $x \in C$, $c \in
\rat$ is a rational constant and $\delta,\delta_1,\delta_2 \in
\Phi(X)$.

A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
where: $\Sigma$ is a finite set of input events, $S$ is a finite set
of states, $S_0, F \subseteq S$ are sets of initial and final states,
respectively, $C$ is a finite set of clocks and $E \subseteq S \times
\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
\item $s_0 \in S_0$ and $\gamma_0(x)$ for all $x \in C$, 
%
\item for all $i \in [n]$, there exists a transition
  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
\end{compactitem}
The run is accepting iff $s_n \in F$, in which case $T$ accepts
$w$. As usual, we denote by $\lang{T}$ the set of finite words
accepted by $T$. It is well-known \cite{AlurDill94} that, in general,
there is no timed automaton accepting the complement language
$(\Sigma\times\real)^* \setminus \lang{T}$ and, moreover, the language
inclusion problem is undecidable.

Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
first-order alternating automaton $\A_T =
\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
variable $t$, ranging over $\real$, such that each timed word $w =
(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
  w\in\lang{T}}$. The only difficulty here is capturing the fact that
all the clocks of $T$ evolve at the same pace, which is easily done
using a technique from \cite{Fribourg98}, that consists in replacing
each clock $x_i$ of $T$ by a variable $y_i$ who tracks the difference
between the values of $t$ and $x_i$, instead of the value of $x_i$
directly.

Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
$\Delta_T$ contains the rule:
\[\begin{array}{rcl}
q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} \exists y'_1 \ldots \exists y'_k
~.~ t > z \wedge \delta(z-y_1,\ldots,z-y_k) & \wedge & 
 \bigwedge_{x_i \in \lambda} y'_i = z \wedge \bigwedge_{x_i \not\in \lambda} y'_i = y_i \\ 
& \wedge & q_j(y'_1,\ldots,y'_k,t)
\end{array}\]
Moreover, nothing else is in $\Delta_T$. We establish the following
connection between a timed automaton and its corresponding first-order
alternating automaton.

\begin{proposition}\label{prop:timed-alternating}
  Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, the
  first-order alternating automaton $\A_T =
  \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$ recognizes the
  language $\lang{\A_T} = \set{d(w) \mid w \in \lang{T}}$. 
\end{proposition}
\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
  \lang{\A_T}$ be a data word. We show the existence of a timed word
  $(a_1,\tau_1) \ldots (a_n,\tau_n) \in \lang{T}$ such that
  $\nu_i(t)=\tau_i$, for all $i \in [1,n]$, by induction on
  $n\geq0$. In fact we shall prove the following stronger
  statements: \begin{compactenum}
  \item each execution of $\A_T$ over $w$ starting with a cube $c \in
    \cube{\minsem{\iota_T}}$ is a linear tree, in which each node has
    at most one child.
    %
  \item for each execution $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
    q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ of $\A_T$, $T$ has an
    execution $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ over
    the timed word $(a_1, \tau_1) \ldots (a_n, \tau_n)$, such that,
    for all $i \in [1,n]$ and all $\ell \in [1,k]$, we have
    $\gamma_i(x_\ell) = \tau_{i-1} - d^i_\ell$.
  \end{compactenum}
  The first point above is by inspection of $\iota_T = \bigvee_{s_i
    \in S_0} q_i(0,\ldots,0)$ and of the rules from
  $\Delta_T$. Indeed, each minimal model of $\iota_T$ corresponds to a
  cube $q(0,\ldots,0)$ and each rule has exactly one predicate atom on
  its right-hand side, thus each node of the execution will have at
  most one successor. The second point is by induction on $n\geq0$.

  ``$\supseteq$'' Let $w = (a_1,\tau_1) \ldots (a_n,\tau_n) \in
  \lang{T}$ be a timed word. By induction on $n\geq0$, we show that
  for each run $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ of $T$
  over $w$, $\A_T$ has a linear execution
  $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
  q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ such that, for all $i \in [1,n]$
  and all $\ell \in [1,k]$, we have $\gamma_i(x_\ell) = \tau_{i-1} -
  d^i_\ell$. \qed}

An easy consequence is that the timed language inclusion problem
``given timed automata $T_1$ and $T_2$, does $\lang{T_1} \subseteq
\lang{T_2}$ ?'' is reduced in polynomial time to the emptiness problem
$\lang{\A_{T_1}} \cap \lang{\overline{\A_{T_2}}} = \emptyset$, for
which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe
further that the transition quantifiers from the encoding of timed
automata into first-order alternating automata can be easily
eliminated by substitution as in $\exists x ~.~ x = t \wedge \phi
\equiv \phi[t/x]$.

\subsection{Register Automata}
\label{sec:ra}

Finite-memory automata, most commonly referred to as register automata
\cite{KaminskiFrancez94} are among the first attempts at lifting the
finite alphabet restriction of classical Rabin-Scott automata. In a
nutshell, a register automaton is a finite-state automaton equipped
with a finite set of registers $x_1,\ldots,x_r$ able to copy input
values and compare them with later the input. Consequently, basic
results from classical automata theory, such as the pumping lemma or
the closure under complement do not hold in this model and inclusion
of languages recognized by register automata is undecidable
\cite{NevenSchwentickVianu03}.

Let $\Sigma$ be an infinite alphabet, $\#$ be a symbol not in $\Sigma$
and $r>0$ be an integer constant. An \emph{assignment} is a word
$\vec{v} = v_1 \ldots v_r$ such that if $v_i = v_j$ and $i \neq j$
then $v_i = \#$, for all $i,j \in [1,r]$. We write $[\vec{v}]$ for the
set $\set{v_i \mid i \in [1,r]}$. A \emph{finite-memory (register)
  automaton} is a tuple $R = \tuple{S,q_0,\vec{u}, \rho, \mu, F}$,
where $S$ is a finite set of states, $q_0 \in S$ is the initial state,
$\vec{u} = u_1 \ldots u_r$ is the initial assignment, $\rho : S
\rightarrow [1,r]$ is the reassignment partial function, $\mu
\subseteq S \times [1,r] \times S$ is the transition relation and $F
\subseteq S$ is the set of final states. A run of $A$ over an input
word $a_1 \ldots a_n \in \Sigma^*$ is a sequence $(s_0,\vec{v}_0)
\ldots (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for all
$i \in [1,n]$, exactly one of the following holds: \begin{compactitem}
\item if there exists $k\in[1,r]$ such that $a_i=(\vec{v}_{i-1})_k$
  then $\vec{v}_i = \vec{v}_{i-1}$ and $(s_{i-1},k,s_i) \in \mu$, 
%
\item otherwise $a_i\not\in[\vec{v}_{i-1}]$, $\rho(s_{i-1})$ is
  defined, $(\vec{v}_i)_{\rho(s_{i-1})} = a_i$, for each $k \in
  [1,r] \setminus \set{\rho(s_{i-1})}$, we have $(\vec{v}_i)_k =
  (\vec{v}_{i-1})_k$ and $(s_{i-1},\rho(s_{i-1}),s_i) \in \mu$.
\end{compactitem}
Intuitively, if the input symbol is already stored in some register,
the automaton moves to the next state if, moreover, the transition
relation allows it, otherwise it copies the input to the register
indicated by the reassignment, erasing its the previous value, and
moves according to the transition relation.

The translation of register automata to first-order alternating
automata is quite natural, because registers can be encoded as
arguments of predicate atoms. Formally, given a register automaton $R
= \tuple{S,s_0,\vec{u},\rho,\mu,F}$, such that $S =
\set{s_0,\ldots,s_m}$, we define the alternating automaton $\A_R =
\tuple{\set{\bullet},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
$\bullet \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
$\#(q_i) = r$ for all $i \in [m]$, $\iota_R \isdef q_0(\vec{u})$, $F_R
\isdef \set{q_i \mid s_i \in F}$ and, for each transition $(s_i, k,
s_j) \in \mu$, $\Delta_T$ contains the rule:
\[q_i(y_1,\ldots,y_r) \arrow{a(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
\bigwedge_{i=1}^r x \neq y_i \wedge q_j(y_1,\ldots,y_{k-1},x,y_{k+1},
\ldots, y_r)\] Moreover, nothing else is in $\Delta_R$. The connection
between register automata and first-order alternating automata is
stated below.

\begin{proposition}\label{prop:register-alternating}
  Given a register automaton $R = \tuple{S,s_0,\vec{u},\rho,\mu,F}$
  over an infinite alphabet $\Sigma$, the first-order alternating
  automaton $\A_R = \tuple{\set{\bullet},Q_R,\iota_R,F_R,\Delta_R}$
  recognizes the languge $\lang{\A_R} = \set{ (\bullet,a_1) \ldots
    (\bullet,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
\end{proposition}
\proof{ ``$\subseteq$'' Let $w = (\bullet,a_1) \ldots (\bullet,a_n)
  \in \lang{\A_R}$. First, it is easy to show that each execution of
  $\A_R$, that starts in some cube $c \in \cube{\minsem{\iota_R}}$, is
  a linear tree with labels $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$
  such that $\vec{v}_0 = \vec{u}$. Second by induction on $n\geq0$, we
  prove that $\A_R$ has a run as above over $w$ only if $R$ has a run
  $(q_0,\vec{v}_0), \ldots, (q_n,\vec{v}_n)$ over $a_1 \ldots
  a_n$. ``$\supseteq$'' Let $w = a_1 \ldots a_n \in \lang{R}$ and
  $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$ be a run of $R$ over $w$,
  such that $\vec{v}_0 = \vec{u}$. By induction on $n\geq0$, we can
  build an execution of $\A_R$ over $(\bullet,a_1) \ldots
  (\bullet,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
  \ldots, q_n(\vec{v}_n)$. \qed}

Consequently, the language inclusion problem ``given register automata
$R_1$ and $R_2$, does $\lang{R_1} \subseteq \lang{R_2}$?'' is reduced
in polynomial time to emptiness problem $\lang{\A_{R_1}} \cap
\lang{\overline{\A_{R_2}}} = \emptyset$, for which
(\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
the encoding of register automata uses no transition quantifiers.

\subsection{Predicate Automata}
\label{sec:pa}

The model of \emph{predicate automata} \cite{Farzan15,Farzan16} has
emerged recently as a tool for checking safety and liveness properties
of parameterized concurrent programs, in which there is an unbounded
number of replicated threads that communicate via global
variables. Predicate automata recognize finite sequences of actions
that are pairs $(\sigma,i)$, where $\sigma$ is from a finite set
$\Sigma$ of program statements and $i \in \nat$ ranges over an
unbounded set of thread identifiers. To avoid clutter, we shall view a
pair $(\sigma,i)$ as a data symbol $(\sigma,\nu)$ where $\nu(x)=i$,
for a designated input variable $x$.

Since thread identifiers can only be compared for equality, the data
theory of predicate automata is the first-order theory of
equality. Moreover, transition quantifiers are only needed for
checking termination and, generally, liveness properties
\cite{Farzan16}. Thus, syntactically speaking, predicate automata are
a subset of first-order alternating automata.

However, the execution semantics of predicate automata differs from
that of first-order automata with respect to the following detail:
initial configurations and successors of predicate automata are
defined using the entire sets of models of the initial sentence and
transition rules, not just the minimal ones, as in our case. Formally,
a run of a predicate automaton $P =
\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\mu$, for each rule
$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
$\mu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
for all $q \in Q \setminus F$.

In fact, as shown next, this more simple execution semantics is
equivalent, from the language point of view, with the semantics given
by Definitions \ref{def:execution} and \ref{def:accepting}. However,
the semantics of first-order alternating automata based on minimal
models is important for its relation to the classical textbook
semantics of boolean alternating automata
\cite{ChandraKozenStockmeyer81}.

\begin{proposition}
  Given a predicate automaton $P =
  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
  first-order alternating automaton that has the same description as
  $P$. Then $\lang{P} = \lang{\A_P}$. 
\end{proposition}
\proof{

\qed}

As before, this result enables using the semi-algorithm of
(\S\ref{sec:emptiness}) for checking emptiness of predicate automata.
We point out that, although quantifier-free predicate automata with
predicates of arity one are decidable for emptiness \cite{Farzan15},
currently there is no method for checking emptiness of predicate
automata with \begin{inparaenum}[(i)]
\item predicates of arity greater than one, or 
\item transition quantifiers. 
\end{inparaenum}
